---
title: "Human Genomics"
subtitle: "Session 1: Introduction to R"
author:
  - name: "Palle Duun Rohde"
    url: https://pdrohde.github.io/
    affiliation: Genomic Medicine, Aalborg University
    email: palledr@hst.aau.dk
date: "2025-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/fs61oo/Documents/GitHub/AAU-human-genomics/")
library(kableExtra)
```

This section provides a brief introduction to fundamental <tt>R</tt> operations. Follow each step carefully, and ensure you understand what each command does and why it works as expected.

#### Calculations
You can use <tt>R</tt> as a powerful calculator to perform standard arithmetic operations using the following mathematical operators:
```{r}
1 + 3   # Addition
9 - 3   # Subtraction
4 * 9   # Multiplication
20 / 5  # Division
```

<tt>R</tt> follows the conventional mathematical order of operations:

-  Parentheses,
-  Exponents,
-  Multiplication/Division,
-  Addition/Subtraction.
	
```{r}
4 / 2^2 + 2 * 2
```

Explanation:

-	The exponent is calculated first: `2^2 = 4`.
-	Then, division and multiplication are performed from left to right: `4 / 4 = 1` and `2 * 2 = 4`.
-	Finally, addition: `1 + 4 = 5`.

Thus, the result of the expression is `5`.

You can modify the default calculation order by using parentheses to explicitly define how the operations should be performed:

```{r}
(((4/2)^2)+2)*2
```

Breakdown of the calculation:

-	 `4 / 2 = 2` ,
-	 `2^2 = 4` ,
-	 `4 + 2 = 6` ,
-	 `6 * 2 = 12` .

<br />		

#### Assigning objects to variables

In <tt>R</tt>, an object is a fundamental data structure that consists of two key components: **data** (the actual values stored) and a **name** (an identifier used to reference the data). Essentially, objects allow you to store, manipulate, and retrieve data efficiently within your <tt>R</tt> environment.

For example, you can create an object that holds the numbers from 1 to 5 and assign it a name for easy reference:
```{r}
my_numbers <- c(1, 2, 3, 4, 5)
```

In this example:

-  `my_numbers` is the **name** of the object.
- ` c(1, 2, 3, 4, 5)` is the **data**, representing a sequence of numbers.

Once the object is created, you can reference it by its name to access or manipulate the stored data:
```{r}
print(my_numbers)  # Outputs: 1 2 3 4 5
```

To create an object in In <tt>R</tt>, you use the **assignment operator** `<-` (_preferred_) or `=` to assign values to a named variable. For example:
```{r}
x <- 10  # Assigns the value 10 to the object named 'x'
y = 20   # Assigns the value 20 to the object named 'y' (alternative syntax)
```

The arrow (`<-`) is generally preferred for several reasons related to readability, convention, and potential confusion with function arguments.

1.  **Readability and Clarity**
    -  The arrow (`<-`) visually indicates the flow of data from right to left, making it clear that a value is being assigned to an object.  
2.  **Consistency with R Language Conventions**
    -  The arrow (`<-`) is the standard assignment operator in <tt>R</tt> and is used consistently in official documentation, tutorials, and best practices.
    -  Experienced <tt>R</tt> programmers and packages predominantly use `<-`, making it the conventional and expected choice.
    -  Following this convention ensures consistency and avoids confusion when collaborating with others.
3.  **Avoiding Confusion with Function Arguments**
    -  In function calls, `=` is used to specify argument values. If used for assignment, it can lead to unintended errors or confusion.
4.  **Flexibility and Pipe Operator Compatibility**
    -  The tidyverse ecosystem (e.g., `dplyr` and `ggplot2`) uses `%>%` (the pipe operator) heavily, and `<-` provides better compatibility and readability in such pipelines, e.g., `data <- mtcars %>% filter(mpg > 20)`

<tt>R</tt> provides several types of objects to store different kinds of data, including:
-  **Vectors** – A sequence of elements of the same type (e.g., numeric, character).
-  **Matrices** – A two-dimensional collection of elements.
-  **Data Frames** – A table-like structure where columns can hold different data types.
-  **Lists** – A flexible container that can hold multiple types of elements.

```{r}
vec <- c(1, 2, 3)               # Vector
mat <- matrix(1:9, nrow=3)      # Matrix
df <- data.frame(a=1:3, b=4:6)  # Data Frame
lst <- list(name="Alice", age=25, scores=c(80, 90, 85))  # List
```

Once an object is created, you can:

-  View its contents using functions like `print()` or simply typing its name.
-  Modify it by assigning new values.
-  Remove it from memory using `rm(object_name)`.

Example of modifying an object:
```{r}
my_numbers <- c(10, 20, 30)  # Redefining the object with new values
rm(my_numbers)               # Deleting the object
```

In <tt>R</tt>, the same result can often be achieved using multiple approaches. A common example is generating a vector containing the numbers from 1 to 10. R provides several ways to accomplish this, each with its own advantages in terms of readability and flexibility.

The `c()` function, which stands for _combine_, allows you to manually create a vector by specifying each individual number. This method provides explicit control over the elements in the vector but can be cumbersome for long sequences.

```{r}
c(1,2,3,4,5,6,7,8,9,10)
```

The `:` operator provides a concise way to generate a sequence of consecutive integers by specifying the starting and ending values.
```{r}
1:10
```

The `seq()` function provides greater flexibility for generating sequences by allowing customization of the start, end, and step size.
```{r}
seq(from=1, to=10, by=1)
```

You can adjust the step size to create sequences with increments other than 1, such as:
```{r}
seq(from = 1, to = 10, by = 2)
```

You can then do operations on the entire vector at once. For example, multiple by 2.
```{r}
x <- 1:10

x*2
```
<br />
	
#### Introduction to Object Classes

In <tt>R</tt>, data is stored in objects, and each object belongs to a specific class, which determines how the data is structured and what operations can be performed on it. Understanding object classes is crucial for efficient data manipulation and analysis in R. The different object types in <tt>R</tt> have unique ways to access their elements. Understanding these indexing methods allows efficient data retrieval and modification. 
<br />
    
##### Vectors

A vector is the most basic data structure in R, representing a sequence of elements of the same type. Vectors can hold numeric, character, logical, or complex values.
```{r, eval=F}
numeric_vec <- c(1, 2, 3.5, 4)
class(numeric_vec)

char_vec <- c("apple", "banana", "cherry")
class(char_vec)  # "character"
```
   
Vectors are one-dimensional sequences of elements, and elements can be accessed using square brackets `[ ]` with an index number.
```{r, eval=F}
v <- c(10, 20, 30, 40, 50)

# Access the second element
v[2]  # 20

# Access multiple elements
v[c(1, 3, 5)]  # 10 30 50

# Access elements using logical indexing
v[v > 25]  # 30 40 50

# Exclude elements using negative indexing
v[-3]  # 10 20 40 50
```
<br />
    
##### Factors

Factors are categorical variables used to represent discrete data with a fixed number of possible values (levels). Factors are useful for handling data that falls into distinct categories, such as gender, colors, or responses in surveys.
```{r, eval=F}
colors <- factor(c("red", "blue", "red", "green"))
class(colors)  # "factor"
levels(colors)  # "blue" "green" "red"
```
   
You can manually specify the levels and their order when creating a factor. 
```{r}
education <- factor(c("High School", "Bachelor", "Master", "PhD", "Bachelor"),
                    levels = c("High School", "Bachelor", "Master", "PhD"))

# Print the factor and check its levels
print(education)
levels(education)
```

When the categories have a logical order (e.g., low to high), you can create ordered factors using the `ordered = TRUE` argument.
```{r}
satisfaction <- factor(c("Low", "Medium", "High", "Medium", "Low"),
                       levels = c("Low", "Medium", "High"),
                       ordered = TRUE)

print(satisfaction)
```
<br />
    
##### Matrices

A matrix is a two-dimensional array where all elements must be of the same type.
```{r, eval=F}
matrix_obj <- matrix(1:9, nrow=3, ncol=3)
class(matrix_obj)  # "matrix"
```

Elements in a matrix can be accessed using row and column indices within `[ , ]`. Use `[row, column]` to access specific values.
```{r}
m <- matrix(1:9, nrow = 3, ncol = 3)

# Access element in the 2nd row, 3rd column
m[2, 3]  # 6

# Access entire 1st row
m[1, ]  # 1 4 7

# Access entire 2nd column
m[, 2]  # 4 5 6
```
<br />
    
##### Data frames

A data frame is a table-like structure where each column can contain different types of data (numeric, character, etc.). It is one of the most commonly used data structures for handling datasets in R.
```{r}
df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30), Score = c(95.5, 89))
class(df)  # "data.frame"
```
    
Data frames are similar to tables and can be indexed using row and column indices, column names, row names, or the `$` operator.
```{r}
df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30), Score = c(95, 85))

# Access a single column by name
df$Age  # 25 30

# Access the first row, second column
df[1, 2]  # 25

# Access all rows for 'Score' column
df[, "Score"]  # 95 85

# Filter rows based on condition
df[df$Age > 26, ]  # Returns Bob's data
```
<br />
    
##### Lists

A list is a flexible data structure that can store elements of different types, including vectors, matrices, data frames, and even other lists.
```{r, eval=F}
my_list <- list(name = "Alice", age = 25, scores = c(90, 85, 88))
class(my_list)  # "list"
```

Elements in lists are accessed using double square brackets `[[ ]]` or the dollar sign `$`.
```{r}
lst <- list(name = "Alice", age = 25, scores = c(90, 80, 85))

# Access by index
lst[[1]]  # "Alice"

# Access by name
lst$age  # 25

# Access nested elements
lst$scores[2]  # 80
```
<br />
    
##### Arrays

Arrays are similar to matrices but can have more than two dimensions. They are used for multi-dimensional data storage.
```{r, eval=F}
array_obj <- array(1:12, dim = c(2, 3, 2))
class(array_obj)  # "array"
```
    
Arrays are multi-dimensional objects, and elements can be accessed using multiple indices within `[ , , ]`.
```{r}
a <- array(1:12, dim = c(2, 3, 2))

# Access element from the 1st row, 2nd column, 1st matrix
a[1, 2, 1]  # 3

# Access the entire 2nd matrix
a[ , , 2]
```

<br />

#### Introduction to Mathematical Operations on Matrices 

Just like vectors, matrices in R allow for straightforward mathematical operations. Since matrices are essentially two-dimensional arrays of numbers, you can perform various element-wise and matrix-specific operations easily.


<br />
<br />	
    
    
    
    
    
    
    
    
    
    
    
    